---
title:  a:class <- b
author: "Romain François"
date:  2014-07-28
tags: ["R", "grammar", "beepr"]
banner: "img/banners/static.png"
---

Usually in strongly typed languages, like C++, the type of a variable comes before the variable, e.g.:

```
int x = 23 ;  
```

So  `x` is declared of type `int` and assign the value `23`.

Now, some languages do things differently, e.g. in julia:

```
x::Int8 = 1000  
```

or go :

```
var i int = 1 ;  
```

So here is a curious thing we can do with R:

```{r eval=FALSE}
`:<-` <- function(x, y, value){
    cl <- deparse(substitute(y))
    target <- deparse(substitute(x))
    if( !is(value, cl) ) {
        beepr::beep(7)
        stop(sprintf("incompatible, expecting %s", cl ) )
    }
    assign( target, value, parent.frame() )
}
```

The idea is that we can do something like this:

```
x  :integer <- 3L  
x  :integer <- "foo"  
```

It does not work if `x` does not already exist, which makes this kind of useless:

```{r eval=FALSE}
> x :integer <;- 3
Erreur dans x:integer <- 3 : objet 'x' introuvable  
```

However, if `x` already exist, it does:

```
> x <- NULL
> x :integer <- 3L
> x
[1] 3
> x :integer <- "foo"
Erreur dans `:<-`(`*tmp*`, integer, value = "foo") :  
  incompatible, expecting integer
```

This is not particularly useful. For it to be more useful, we would need the R grammar to recognize `a:b <- c` and do something meaningful with it.
