---
title: Rows don't exist
author: Romain François
date: '2017-10-20'
slug: rows-don-t-exist
categories: []
tags:
  - tidyverse
  - Rcpp
  - naniar
  - dplyr
draft: true
---

```{r include=FALSE}
library(tibble)
library(dplyr)
```


Every now and then, I get questions about row-wise operations in the `#tidyverse`, and most of the time I 
end up coming up with a solution that is column-wise instead of row-wise. 

Let's pretend we don't know about 
`rowSums` and we want to count the number of values greater than 2 in each row of this tibble based on the 
columns `x`, `y` and `z`

```{r}
d <- tribble( 
  ~x, ~y, ~z, ~group, 
  1, 2, 3, "a", 
  2, 3, 2, "b", 
  1, 2, 3, "a", 
  3, 1, 2, "a"
)
```

It is easy enough to want to use `apply(, 1)` for this task, i.e. first `select_if` the 
columns of interest and then apply a function to each `row` 

```{r}
select_if(d, is.numeric) %>% 
  apply( 1, function(x) sum(x>1) )
```

The first problem here is that we escape from the tidyverse and the first thing 
`apply` will do is to create a disgraceful matrix from our carefully crafted tibble.
Of course I'm exagerating, and of course matrices are fine, when they are fine. It's just not one of these
cases. 