---
title: Unicode, utf-8, strings and emojis
author: Romain François
date: '2017-08-03'
slug: unicode-utf-8-strings-and-emojis
categories: []
tags:
  - unicode
  - utf8
  - strings
  - emojis
  - utf8splain
  - uni
  - package
  - R
banner: "img/banners/utf8splain.png"
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

I've been somewhat obsessing about emojis lately, it all started when I wanted to
check which emojis were used [on twitter](http://romain.rbind.io/blog/2017/07/11/emojis-at-user2017/)
during useR this year.

`r htmltools::HTML("{{< tweet 883199852347367424 >}}")`

But this post is not really about emojis, because my [emojitsu](https://github.com/ThinkRstat/emojitsu)
package is not ready yet, but here's a preview anyway.

`r htmltools::HTML("{{< tweet 892375718247706626 >}}")`

So I'll blog specifically about emojis later, but this has led me to digress down
the 🐇 hole, because emojis are made of unicode runes typically encoded into utf-8 strings. Most of the
concepts in that last sentence were quite mysterious to me not so long ago, and I believe we should
collectively know more about unicode and utf-8. I learned some of the basics from
the [Strings, bytes, runes and characters in Go](https://blog.golang.org/strings) post in the go
blog, and [The Absolute Minimum Every Software Developer Absolutely, Positively Must Know About Unicode and Character Sets (No Excuses!)](https://www.joelonsoftware.com/2003/10/08/the-absolute-minimum-every-software-developer-absolutely-positively-must-know-about-unicode-and-character-sets-no-excuses/).

Typically when I want to understand something, I make an R package, I guess I really want
to understand this, as I am making not 1, not 2 but 3 packages (if I count `emojitsu`).

 - `uni` : contains a tibble of unicode runes
 - `utf8splain` : to get to the 0 and 1 of utf-8 string encoding
 - `emojitsu` : grammar of emoji, or at least programmatic manipulation of them.

The world has changed now, and strings can no longer be considered as mere sequences of single
characters (bytes). The `uni::code` tibble contains the `r nrow(uni::code)` unicode runes
(aka code points). btw, the generation of the `uni::code` tibble contains some interesting
tidyverse gymnatics, perhaps I'll write another post about that, but let's not digress more yet.

```{r}
uni::code
```
