---
title: lazy absolute values function with ALTREP and C++
author: Romain François
date: '2018-10-21'
slug: lazy-abs-altrep-cplusplus
categories: []
tags:
  - altrep
  - c++
  - R
---

This is a follow up [to this post about ALTREP and C++](https://purrple.cat/blog/2018/10/14/altrep-and-cpp). 
This covers another part of the [altrepisode](https://github.com/romainfrancois/altrepisode) `r emo::ji("package")`
to showcase another aspect of ALTREP, the `lazy_abs()` function and underlying C++ implementation.

## What do we ALTREP this time

In R, `abs()` is a vectorized function, i.e. given a vector `x` of `n` values it returns another vector of the same 
size that contains absolute values of each element of `x`. 

Here, we create `lazy_abs()` that does the same thing but lazily, i.e. it only creates the vector 
if absolutely `r emo::ji("rolling_on_the_floor_laughing")` needed. 

One immediate consequence, is that `lazy_abs()` is much faster, because at first it does not 
do anything

```{r}
# devtools::install_github("romainfrancois/altrepisode")
library(altrepisode)
x <- rnorm(1e5)
identical(abs(x), lazy_abs(x))
bench::mark(abs(x), lazy_abs(x))
```

Depending on what you do with the result of `lazy_abs()` it might be materialized, i.e. 
use its own contiguous memory, or not. 

## Implementation

Remember from [this previous post](https://purrple.cat/blog/2018/10/14/altrep-and-cpp) that 
we structure an ALTREP object using two arbitrary R objects `data1` and `data2`. Our lazy 
abs object uses both. 

 - `data1` is used to keep a reference of the input. We carefully mark it as immutable so that
   according to R's copy on modify semantics, if some other user of `data1` wants to modify it, 
   a copy has to be generated.
 - `data2` holds the result, once it has been materialized, but initially is set to `NULL`

The `inspect` implementation can help us illustrate the lazy materialization property: 

```{r}
y <- lazy_abs(rnorm(5))
.Internal(inspect(y))

# head does not call DATAPTR
head(y, 2)
.Internal(inspect(y))

# but mean does
mean(y)
.Internal(inspect(y))
```

## Construction

The `lazy_abs::Make` function implements the strategy described above. We need
the `MARK_NOT_MUTABLE` macro to (the name says it all), and the `R_new_altrep`
function to create an instance of our `lazy_abs::class_t` class. 

```cpp
// make a new altrep object of class lazy_abs::class_t
//
// the object uses both altrep data
// - data1: the original numeric vector we want to abs lazily
// - data2: NULL if the object has not been materialized yet, the entire abs(data1) otherwise
static SEXP Make(SEXP data){
  // so that other holders of data know they have to make a copy
  // when they mutate it
  MARK_NOT_MUTABLE(data);

  // data2 is originally set to NULL
  return R_new_altrep(class_t, data, R_NilValue);
}
```

## Methods

Before we look at the details of the materialization of the object, let's review the various
ALTREP methods. 

The number of elements is simply the number of elements of the input:

```cpp
static R_xlen_t Length(SEXP vec){
  return LENGTH(R_altrep_data1(vec));
}
```

The `Dataptr_or_null` returns a pointer to the start of the results memory if it has already been 
materialized, and NULL otherwise. Users of this object can take the NULL result as a signal 
that the object might be expensive to materialize and prefers individual access if possible. 

```cpp
static const void* Dataptr_or_null(SEXP vec){
  SEXP data2 = R_altrep_data2(vec);
  if (data2 == R_NilValue) return nullptr;

  return STDVEC_DATAPTR(data2);
}
```

The `Dataptr` returns the pointer to the start of the memory, no matter what. Its implementation 
delegates the work to `Materialize` which we'll review later: 

```cpp
static void* Dataptr(SEXP vec, Rboolean writeable){
  return STDVEC_DATAPTR(Materialize(vec));
}  
```

Accessing the `i`th element can use the lazyness, if the result has been materialized, 
just return the `i`th element of the materialized result, otherwise return the 
`abs`olute value of the `i`th element of the input: 

```cpp
static double real_Elt(SEXP vec, R_xlen_t i){
  SEXP data2 = R_altrep_data2(vec);
  if (data2 == R_NilValue) {
    SEXP data1 = R_altrep_data1(vec);
    return ::fabs(REAL_ELT(data1, i));
  } else {
    return static_cast<double*>(STDVEC_DATAPTR(data2))[i];
  }
}
```

There are some subtleties in this implementation of `real_Elt`: 

 - in the non materialized branch, we use the `REAL_ELT` macro. This
   is an ALTREP aware macro that will call the appropriate `real_Elt` 
   method if `data1` is itself some ALTREP object. 
   
 - in the materialized branch, because we are in charge of `data2` and we 
   know (trust me for now, we'll see that in a minute) that it is a standard 
   R vector (i.e. not an ALTREP) we can directly use the `STDVEC_DATAPTR`
   macro to get to the start of the memory. We don't have to pay `r emo::ji("dollar")`
   for the "is it an ALTREP object" test. 
   
I'll skip the [Inspect](https://github.com/romainfrancois/altrepisode/blob/1511d14b17ab4999b48fb4be9a9e493ebbbcdd9c/src/lazy_abs.cpp#L74)
and [Get_region](https://github.com/romainfrancois/altrepisode/blob/1511d14b17ab4999b48fb4be9a9e493ebbbcdd9c/src/lazy_abs.cpp#L127)
methods in this post. 

## Materialization

The critical part is the implementation of the materialization. 

```cpp
// materialize data2 if needed and return it
static SEXP Materialize(SEXP vec) {

  SEXP data2 = R_altrep_data2(vec);
  if (data2 == R_NilValue) {
    // so we do have to materialize
    SEXP data1 = R_altrep_data1(vec);

    // allocate a standard numeric vector for data2
    R_xlen_t n = LENGTH(data1);
    data2 = PROTECT(Rf_allocVector(REALSXP, n));
    // we know it is a standard vector, so it's
    // fine to get its pointer
    double* p_data2 = REAL0(data2);

    // we need to treat the data differently depending
    // on its altrep properties
    auto p_data1 = DATAPTR_OR_NULL(data1);
    if (p_data1){
      // the data from data1 is contiguous, we can scan the values
      // from data1, apply the fabs function and fill data2
      auto p = static_cast<const double*>(p_data1);
      std::transform(
        p, p + n, // the input range
        p_data2,  // the output start
        [](double x){ return fabs(x) ; } // We can't just use fabs because there are overloads.
      );
    } else {
      // the data is not contiguous, so we avoid materializing it
      // and rather go through it element by element
      for (R_xlen_t i=0; i<n; i++, ++p_data2) {
        *p_data2 = ::fabs(ALTREAL_ELT(data1, i)) ;
      }
    }
    
    // keep data2 and release its protection
    R_set_altrep_data2(vec, data2);
    UNPROTECT(1);
  }
  return data2;
}
```

There are a few cases we need to consider for an ALTREP-aware behaviour.

If `data2` is already not NULL, it means it has been already materialized, so we just return it, that's the easy case. 

Otherwise, we start by allocating memory for `data2` with `Rf_allocVector`, we need to protect
this from the garbage collector (actually we don't because currently the gc is disabled during 
ALTREP `Dataptr` calls, but there's no guarantee that this will always be the case). 

```cpp
// allocate a standard numeric vector for data2
R_xlen_t n = LENGTH(data1);
data2 = PROTECT(Rf_allocVector(REALSXP, n));
```

We then need to access the contigious memory of the object. since 
we've just generated it, we know it's a standard R vector, so we can use 
the `REAL0` macro: 

```
// we know it is a standard vector, so it's
// fine to get its pointer
double* p_data2 = REAL0(data2);
```

Great, we have the start of the memory `p_data2` and a size `n` so we 
can now fill the data with absolute values of the input. But we still have two cases
to consider because we don't know the altrep-ness or `data1`. 

We call the `DATAPTR_OR_NULL` macro to either retrieve the data pointer of the input if
already available, i.e. if it's a standard R vector or another ALTREP that can access
its data pointer without memory allocation. 

```cpp
auto p_data1 = DATAPTR_OR_NULL(data1);
```

If we did get a non null pointer, we can reliably traverse the entire memory
of `data1`, call `::fabs` on each element and store that into `data2`, the 
C++ algorithm `std::transform` is perfect for the task. 

```cpp
auto p = static_cast<const double*>(p_data1);
std::transform(
  p, p + n, // the input range
  p_data2,  // the output start
  [](double x){ return fabs(x) ; } // We can't just use fabs because there are overloads.
);
```

If we did get a null pointer, it means that `data1` is not available as a contiguous memory
without allocation, so we go through elements of `data1` one by one using the `ALTREAL_ELT`
macro which will dispatch to the appropriate `real_Elt` method. 

```cpp
// the data is not contiguous, so we avoid materializing it
// and rather go through it element by element
for (R_xlen_t i=0; i<n; i++, ++p_data2) {
  *p_data2 = ::fabs(ALTREAL_ELT(data1, i)) ;
}
```

Finally, once all the absolute values are in `data2`, we can set it as the `data2` of our
ALTREP object, and remove its protection, because the altrep object now protects it. 

```cpp
R_set_altrep_data2(vec, data2);
UNPROTECT(1);
```

## ALTREP composition

Because of the way `Materialize` is implemented, the `lazy_abs()` respects the 
ALTREP-ness of the input object, so they can be composed. For example, we can 
take a `lazy_abs()` of the `stdvec_doubles` class we discussed in 
[the previous post](https://purrple.cat/blog/2018/10/14/altrep-and-cpp/)

```{r}
x <- doubles()
y <- lazy_abs(x)
```

At first `y` is a non materialized `lazy_abs` ALTREP object that keeps a reference of 
a `stdvec_doubles` ALTREP object: 

```{r}
.Internal(inspect(y))
```

If we change one value in `x`, it does not affect `y` because we've marked it as non mutable: 

```{r}
x[1] <- 7
.Internal(inspect(x))
```

`x` is now a standard numeric vector, but the `data1` of `y` is still the same 
ALTREP powered `stdvec_doubles` object. 

```{r}
.Internal(inspect(y))
```

Then if we force materialization of `y` using any R function that happens to 
need contiguous memory, then `data2` is made

```{r}
z <- sort(y)
z
.Internal(inspect(y))
```

## Next time

We'll have a look at ALTREP vectors of strings, although I haven't written the code for it 
yet in [altrepisode](https://github.com/romainfrancois/altrepisode) so maybe not `r emo::ji("shrug")`. 

