---
title: Strings know their own length
author: Romain François
date: '2018-03-05'
slug: strings-know-their-own-length
categories: []
tags:
  - R
  - C
  - C++
  - rapi
---

```{r, include=FALSE}
# knitr::knit_hooks$set( document = emo::ji_glue )
```

## Strings are arrays of bytes

Strings are *hard*, especially when you start to add concepts like encoding and 
locales, etc ... but that's not what we are talking about here. Let's `r emo::ji("rewind")`. 

Strings are *simple*, they are just a consecutive sequences of characters, let's keep 
hiding a lot under the carpet `r emo::ji("speak_no_evil")`, 
strings are just a (null terminated) array of bytes. The length of the string is the number of
non null bytes. 

A simple way to get that information in C would be a loop (it's ok to write loops in C). 

```c
int string_length( const char* s){
  int n = 0 ;
  for( char* p = s; *p != 0; p++) n++ ;
  return n ;
}
string_length( "head, shoulders, knees and toes") ;
```

This just counts the number of bytes (aka `char`) in the string, obviously in real life
what a character is is more complicated than a `byte`<span class="note">some characters, like say emojis are 
made of several bytes</span>, but we said this was another day's problem. 

We don't have to write the loop manually because we know about the 
C function `strlen` ;

```c
strlen( "head, shoulders, knees and toes" ) ;
```

## What about R strings

You don't typically see individual strings in R because R has no 
notion of scalars, you see character vectors of length one, perhaps through 
the `Rcpp::CharacterVector` class. 

You don't see them, but they exist internally as one of the SEXP types, namely `CHARSXP`. 
When you have a character vector (a `STRSXP`) and you want a single string (a `CHARSXP`) 
you may use the `STRING_ELT` macro. 

Then if you want the actual string, you can use the `CHAR` macro. 

```{r}
Rcpp::cppFunction( '
int rstring_length_strlen( SEXP sv ){
  SEXP s = STRING_ELT(sv, 0) ;
  return strlen( CHAR(s) ) ;
}
')
s <- "head, shoulders, knees and toes"
rstring_length_strlen( s )
nchar(s)
```

But `CHAR` is not the only thing you can do to a `CHARSXP`, you can also 
use the `LENGTH` macro, which retrieves information that is stored in the `SEXP` 
alonside the actual string, so R strings already know their length. 


```{r}
Rcpp::cppFunction( '
int rstring_length_LENGTH( SEXP sv ){
  SEXP s = STRING_ELT(sv, 0) ;
  return LENGTH( s ) ;
}
')
rstring_length_LENGTH( s )
```

## Back story

This might be an obscure post, it's been on this [blog issues](https://github.com/purrple/website/issues/1) for a while now. I guess
I've not materialised it yet is because of my conception that this is either known
by people for which this relevant, or it is not relevant to other people. Either 
you already know, or you don't have to care. 

I'll try to force myself to write about these types of things when I can identify that 
at some point in the past, I would have benefited from knowing and I did not. 

And that's quite the case, otherwise the `size` method of [`Rcpp::String`](https://github.com/RcppCore/Rcpp/blob/29f28c45c9a9aa33755a79f368f7158de4d2a49a/inst/include/Rcpp/String.h) would use 
`length` when in fact there isn't even a `size` method `r emo::ji("scream")`. 

