---
title: multiple lags with tidy evaluation
author: Romain François
date: '2018-03-02'
slug: multiple-lags-with-tidy-evaluation
categories: []
tags:
  - bangbang
  - tidyeval
  - dplyr
  - purrr
  - rlang
---

```{r, include = FALSE}
library(dplyr)
library(purrr)
library(rlang)
```

## multiple lags

This came up on twitter during [rstudio::conf](https://www.rstudio.com/conference/) as 
a question from [Simon Jackson](https://twitter.com/drsimonj). That's a nice kind of question
because it comes with example code, so I had a go at it, but I had not taken the time to 
promote it to a blog post. Let's fix that. 

`r blogdown::shortcode_html( "tweet", "959717388764803072")`

Simon's [code](https://gist.github.com/drsimonj/2038ff9f9c67063f384f10fac95de566) is using 
the combination of `mutate_at` and the lazyeval style `funs_` to make functions from strings. 

```{r}
d <- data_frame(x = seq_len(100))
lags <- seq(10)
lag_names <- paste("lag", formatC(lags, width = nchar(max(lags)), flag = "0"), 
  sep = "_")
lag_functions <- setNames(paste("dplyr::lag(., ", lags, ")"), lag_names)
lag_names
lag_functions
```

This is perfectly valid, but it has an `eval/parse` feel, and the `funs_` is not 
inline with current tidy evaluation. 

## jetlag

My [first take](https://gist.github.com/romainfrancois/6a2203faacfa8ae849f29f83b56e01da) 
was to create a function, called `jetlag` that would do the 
whole operation using tidy evaluation. <span class="note">The name was probably reflecting my
physical state at the time, as it turns out flying from Paris to San Diego with 
a 6 hours connection in New York makes jetlag. </span>

```{r}
jetlag <- function(data, variable, n=10){
  variable <- enquo(variable)
  
  indices <- seq_len(n)
  quosures <- map( indices, ~quo(lag(!!variable, !!.x)) ) %>% 
    set_names(sprintf("lag_%02d", indices))
  
  mutate( data, !!!quosures )
  
}
jetlag(d, x, 3)
```

Let's break it down in steps. The function takes 3 parameters: 

 - `data` a data frame we want to process
 - `variable` a symbol that corresponds to a column name in `data`
 - `n` the number of lags to create
 
The first thing it does is to pass `variable` to `enquo`. This allows `variable` to be 
passed by expression rather than by value. This is why when we call `jetlag` we
don't have to pass `x` as a string, but just as a symbol. 
 
Next, the function makes a list of quosures  by iterating with `map` over the 
`seq_len(n)` sequence. 

```{r}
# this is the same as `enquo`, but outside of a function
variable <- quo(variable)
indices <- seq_len(3)
quosures <- map( indices, ~quo(lag(!!variable, !!.x)) ) %>% 
  set_names(sprintf("lag_%02d", indices))
```

The first `!!` unquotes `variable`, i.e. replaces `variable` by `x`, the second 
`!!` replaces injects the value of the placeholder `.x`. 

Piping into `set_names` gives names to the list, `mutate` can then use these names. 

```{r}
quosures
names(quosures)
```

Finally, the `!!!` injects the expressions in `mutate`. In essencen, what `jetlag` is doing is
programmatically build something similar to this repetitive expression: 

```{r}
mutate( d, 
  lag_01 = lag(x, 1), 
  lag_02 = lag(x, 2), 
  lag_03 = lag(x, 3)
)
```

but with hopefully with nicer (or at least shorter) syntax: 

```{r}
jetlag(d, x, 3)
```

## lags

As this is often the case, I immediately posted this on twitter. 

`r blogdown::shortcode_html( "tweet", "959811892922347520")`

Only to realise it a few minutes later that this is not as 
reusable/composable as it could be. In fact the interesting
thing from this function is the creation of the list of quosures, not so much the 
actual call to `mutate`. That's an example where doing *less* makes something *more*
useful. 

So my second take was to make it a 
[smaller function](https://gist.github.com/romainfrancois/469ed8224ba4be48befec28cb1e1ff80), called `lags`
that would just make the quosures. Naming functions is something 
I often struggle with, so when it appears as obvious as just using plural, 
I know I'm on the right track. 

This was also an, opportunity to capture the name of the variable the lags 
operate on, and use it as a component to the names of the columns to create. 

```{r}
lags <- function(var, n=10){
  var <- enquo(var)
  
  indices <- seq_len(n)
  map( indices, ~quo(lag(!!var, !!.x)) ) %>% 
    set_names(sprintf("lag_%s_%02d", quo_text(var), indices))
  
}
```

`lags` only needs a name and a number of columns, it does not need `data` as the 
previous attempt. This makes it easier to understand what is going on: 

```{r}
lags( xyz, 4 )
```

`lags` creates the quosures and then we can unquote splice them "manually". 

```{r}
mutate( d, !!!lags(x, 3) )
```

This also makes it slightly easier to use it on use it on multiple variables at once. 


```{r}
d <- data_frame( x = 1:10, y = letters[1:10])
d %>% 
  mutate( !!!lags(x, 3), !!!lags(y,3) )
```

## Back story

Taking the time to write this as a blog post now was inspired by [Mara's](https://twitter.com/dataandme)
promotion of one of my early [blog post](https://purrple.cat/blog/2017/06/28/filter-with-context/)
about tidy evaluation when [dplyr 0.7](https://blog.rstudio.com/2017/06/13/dplyr-0-7-0/) was released. 

`r blogdown::shortcode_html( "tweet", "968973417197309954")`

Tidy evaluation is one of these things that feel natural once you are more familiar with the concepts, 
I hope this example helps. I expect to be posting more of these as I come up with uses cases. 

`r blogdown::shortcode_html( "tweet", "969121867729645568")`





