---
title: ALTREP and C++
author: Romain François
date: '2018-10-14'
slug: altrep-and-cpp
categories: []
tags:
  - R
  - C++
  - altrep
---

Traditionally, R vectors (numeric, ...) are formed of one header ditrectly followed by the actual 
contiguous data. I'll spare the details about the headers because even though I have a fair understanding, 
I don't necessarily want to propagate my misconceptions `r emo::ji("grimacing")`, 
and also because it is mostly irrelevant to this post. 

Since R 3.5.0, the implementation of vectors benefits from `ALTREP` (alternative representation), which 
challenges this and decouples the header with the actual data. This has two main use cases: 

 - when the memory comes from somewhere else (a memory mapped file, an `arrow` array or whatever), i.e.
   some contiguous memory that comes from somewhere. `ALTREP` lets you create an R vector 
   that *points* to that memory. 
   
 - when you only need part of the vector, e.g. a single value. Here `ALTREP` lets you define 
   what is the `i`th value of the vector, without having to materialize it. The canonical 
   example is `for(i in 1:n){}`. You never need the entire vector `1:n`, you only need 
   one value at a time. 
   
As the projects I'm currently involved with (`dplyr` and `arrow`) might ultimately 
benefit from `ALTREP`, I've `r emo::ji("person juggling")` with it in the 
[altrepisode](https://github.com/romainfrancois/altrepisode) `r emo::ji("package")`. 

In this post, I'm covering the first use case with numeric vectors, in particular
creating an R altrep vector that borrows data from a C++ `std::vector<double>`. 

## The Altrep.h header

To manipulate `ALTREP` objects, create your own classes, you need to include 
the `R_ext/Altrep.h` header file. Unfortunately in R 3.5.0, the 
[`Altrep.h`](https://github.com/wch/r-source/blob/trunk/src/include/R_ext/Altrep.h)
is not C++ friendly, so you need to be careful when you include it. The situation has
been [fixed](https://github.com/wch/r-source/commit/14365e4419d08ced8e987d5d49b26843d1324297) 
recently so if you rely on R-devel things are easier. 

In the meantime, here is my workaround: 

```cpp
// to manipulate R objects, aka SEXP
#include <R.h>
#include <Rinternals.h>
#include <Rversion.h>

// because we need to initialize the altrep class
#include <R_ext/Rdynload.h>

#if R_VERSION < R_Version(3, 6, 0)

// workaround because R's <R_ext/Altrep.h> not so conveniently uses `class`
// as a variable name, and C++ is not happy about that
//
// SEXP R_new_altrep(R_altrep_class_t class, SEXP data1, SEXP data2);
//
#define class klass

// Because functions declared in <R_ext/Altrep.h> have C linkage
extern "C" {
  #include <R_ext/Altrep.h>
}

// undo the workaround
#undef class

#else
  #include <R_ext/Altrep.h>
#endif
```

The <200 lines of (fairly documented) C++ code [stdvec_doubles.cpp](https://github.com/romainfrancois/altrepisode/blob/master/src/stdvec_double.cpp)
contains the implementation that is discussed here. This uses `Rcpp` but only for code generation using 
the attributes feature. This is not using any Rcpp classes (e.g. `NumericVector`), which at this point 
would eliminate the benefits of `ALTREP`. 

## Motivation

Let's start by the end of the file to get some motivation. The `doubles` function below creates an 
R altrep object backed by a `std::vector<double>`. 

```cpp
//' an altrep object that wraps a std::vector<double>
//'
//' @export
// [[Rcpp::export]]
SEXP doubles() {
  // create a new std::vector<double>
  //
  // this uses `new` because we want the vector to survive
  // it is deleted when the altrep object is garbage collected
  auto v = new std::vector<double> {-2.0, -1.0, 0.0, 1.0, 2.0};

  // The altrep object owns the std::vector<double>
  return stdvec_double::Make(v, true);
}
```

The details will follow, but for now, let's look at what the object look like
when back on the R side: 

```{r}
library(altrepisode)
x <- doubles()
x
mode(x)
str(x)
```

It looks and feels like any other R numeric vector, that's the point. As far as the R
code is concerned, this is not different than an object that would have been created
by `c`: 

```{r}
y <- c(-2.0, -1.0, 0.0, 1.0, 2.0)
identical(x, y)
```

To see a difference, you have to look at the object with `.Internal(inspect())`, as ALTREP
gives you a way to control how your ALTREP objects are inspected. 

```{r}
.Internal(inspect(x))
.Internal(inspect(y))
```

## Register the ALTREP class

The `stdvec_double::Make` function from the above code chunk creates an R object
of ALTREP class (the ALTREP class is completely orthogonal to 
the R class, as again as far as R code is concerned, nothing has changed). 

For this we need to register an `R_altrep_class_t` object with (in the case
of an altrep `numeric` vector) the `R_make_altreal_class` function. 

This is done at `r emo::ji("package")` initialisation time, thanks 
to the new [`Rcpp::init` attribute](https://github.com/RcppCore/Rcpp/pull/903). 

```cpp
// static initialization of stdvec_double::class_t
R_altrep_class_t stdvec_double::class_t;

// Called the package is loaded (needs Rcpp 0.12.18.3)
// [[Rcpp::init]]
void init_stdvec_double(DllInfo* dll){
  stdvec_double::Init(dll);
}
```

ALTREP is a C api, relying on C functions, but because this is C++, 
I've squashed the functions together as static functions of the `stdvec_double` C++ class, hence
the `stdvec_double::Init` call here. `Init` looks like this: 

```cpp
static void Init(DllInfo* dll){
  class_t = R_make_altreal_class("stdvec_double", "altrepisode", dll);

  // altrep
  R_set_altrep_Length_method(class_t, Length);
  R_set_altrep_Inspect_method(class_t, Inspect);

  // altvec
  R_set_altvec_Dataptr_method(class_t, Dataptr);
  R_set_altvec_Dataptr_or_null_method(class_t, Dataptr_or_null);

  // altreal
  R_set_altreal_Elt_method(class_t, real_Elt);
  R_set_altreal_Get_region_method(class_t, Get_region);
}
```

First, we register the class with `R_make_altreal_class` function, then 
we replace default methods with custom functions that know how to deal with 
out `std::vector<double>`. 

 - `Length` : what is the length of the vector
 - `Inspect` : what happens when we `.Internal(inspect())` the object
 - `Dataptr` : where is the data (more on that later) ?
 - `Dataptr_or_null` : where is the data (but don't look too hard)
 - `real_Elt` : what is the `i`th element ?
 - `Get_region` : A contiguous region of the data
 
This is I believe `r emo::ji("shrug")` the bare minimum. 

In addition to that, the `stdvec_double` class hosts: 

 - `Make` : to construct one such objet from a *pointer to* a `std::vector`. 
 - `Finalize`: to `delete` the object as the proper time if we own it
 - `Ptr` : to get the pointer
 - `Get` : to get a reference to the `std::vector<double>`
 
 ## Construction
 
 The `stdvec_double::Make` function creates the altrep R object backed by the 
 `std::vector<double>: 
 
```cpp
// Make an altrep object of class `stdvec_double::class_t`
static SEXP Make(std::vector<double>* data, bool owner){
  // The std::vector<double> pointer is wrapped into an R external pointer
  //
  // `xp` needs protection because R_new_altrep allocates
  SEXP xp = PROTECT(R_MakeExternalPtr(data, R_NilValue, R_NilValue));

  // If we own the std::vector<double>*, we need to delete it
  // when the R object is being garbage collected
  if (owner) {
    R_RegisterCFinalizerEx(xp, stdvec_double::Finalize, TRUE);
  }

  // make a new altrep object of class `stdvec_double::class_t`
  SEXP res = R_new_altrep(class_t, xp, R_NilValue);

  // xp no longer needs protection, as it has been adopted by `res`
  UNPROTECT(1);
  return res;
}
```
 
Eventually the R object (aka SEXP) is created with the `R_new_altrep` function, 
which takes the altrep class as first argument, and two other 
arbitrary R objects. These two R objects can be later accessed with `R_altrep_data1`
and `R_altrep_data2` and can be just about anything you like. 

Here we use an external pointer, created with `R_MakeExternalPtr`, as `data1` and 
we don't need anything for `data2` so we use `NULL`. If we own the C++ vector, 
as indicated by the `owner` argument, we register a finalizer so that when the 
external pointer (the R object) is garbage collected, the destructor of the 
C++ object is invoked. 

The `Finalize`, `Ptr` and `Get` functions are conveniences that allow us to go from 
the altrep R object to the C++ vector: 

```cpp 
// finalizer for the external pointer
static void Finalize(SEXP xp){
  delete static_cast<std::vector<double>*>(R_ExternalPtrAddr(xp));
}

// get the std::vector<double>* from the altrep object `x`
static std::vector<double>* Ptr(SEXP x) {
  return static_cast<std::vector<double>*>(R_ExternalPtrAddr(R_altrep_data1(x)));
}

// same, but as a reference, for convenience
static std::vector<double>& Get(SEXP vec) {
  return *Ptr(vec) ;
}
```

Given the altrep object `x` we need to first get to its `data1` with `R_altrep_data1` and then 
cast that to the underlying C++ vector with `R_ExternalPtrAddr` and a `static_cast<>`. Once we have 
this, the rest follows naturally. 

## ALTREP methods
 
ALTREP is divided in several layers depending on the type of object we altrep (maybe it's to soon 
for this to be a verb). The first layer is generic and apply to all ALTREP objects. There might 
be other methods, but for this I've implemented the `Length` and `Inspect` : 

```cpp
// The length of the object
static R_xlen_t Length(SEXP vec){
  return Get(vec).size();
}

// What gets printed when .Internal(inspect()) is used
static Rboolean Inspect(SEXP x, int pre, int deep, int pvec, void (*inspect_subtree)(SEXP, int, int, int)){
  Rprintf("std::vector<double> (len=%d, ptr=%p)\n", Length(x), Ptr(x));
  return TRUE;
}
```
 
In the `Length` method, we are given the altreped (still assuming this is a verb) object. From this object, 
we extract the `std::vector<double>&` with `Get` and then simply call `size()` on it. 

The `Inspect` is a bit more involved, let's just skip it `r emo::ji("see_no_evil")`. 

## ALTVEC methods

Then, we have methods that are only relevant to vector type R objects. In this 
implementation, I have defined the `Dataptr` and `Dataptr_or_null` methods: 

```cpp
// The start of the data, i.e. the underlying double* array from the std::vector<double>
//
// This is guaranteed to never allocate (in the R sense)
static const void* Dataptr_or_null(SEXP vec){
  return Get(vec).data();
}

// same in this case, writeable is ignored
static void* Dataptr(SEXP vec, Rboolean writeable){
  return Get(vec).data();
}
```

In this example, they are identical, but it's not necessarily the case for all altrep 
class implementations. 

The difference is that the `Dataptr_or_null` method is guaranteed to not allocate
additional R memory. If you already have access to a contiguous chunk of memory, 
then return that, otherwise return a null pointer, but this should never allocate. 

The `Dataptr` method is the big `r emo::ji("hammer")`. Whatever your class has to do, 
it must return a pointer to contiguous chunk of memory where the data is. We'll illustrate
this better in another post, but in short if you already have that contiguous memory, 
then return it, if not do whatever it takes, allocate if you have to, but eventually 
get me that memory. 

`Dataptr` is what most ALTREP unaware (e.g. the constructor of `Rcpp::NumericVector` or the
`mean` function from base R) code will use. 

## ALTREAL methods

Eventually, methods specific to numeric vectors. 

```cpp
// the element at the index `i`
//
// this does not do bounds checking because that's expensive, so
// the caller must take care of that
static double real_Elt(SEXP vec, R_xlen_t i){
  return Get(vec)[i];
}

// Get a pointer to the region of the data starting at index `i`
// of at most `size` elements.
//
// The return values is the number of elements the region truly is so the caller
// must not go beyond
static R_xlen_t Get_region(SEXP vec, R_xlen_t start, R_xlen_t size, double* out){
  out = Get(vec).data() + start;
  R_xlen_t len = Get(vec).size() - start;
  return len > size ? len : size;
}
```

`real_Elt` gives you the element at offset `i` of the underlying object, here this 
calls the `std::vector<double>::operator[]` but you can imagine situations
where this is implemented differently, e.g. for `1:n` you can imagine this would just 
return `i+1` without having to rely on anything else. 

`Get_region` again is a bit more involved, so I'll skip it too. You can do what I do and guess
what it is supposed to do based on its name. There can be situations where you can have 
access to contiguous memory for part of the vector. 

## What else to read

I'll write some more about this in another post. In the meantime, here are a few pointers: 

 - Luke Tierney's [presentation](http://homepage.divms.uiowa.edu/~luke/talks/uiowa-2018.pdf) about ALTREP. 
 - The [ALTREP Examples](https://github.com/ALTREP-examples) organization on github hosts packages that 
   showcase ALTREP. Most of what I did on [altrepisode](https://github.com/romainfrancois/altrepisode)
   is inspired from the [simplemmap](https://github.com/ALTREP-examples/Rpkg-simplemmap) `r emo::ji("package")`. 
   The [mutable](https://github.com/ALTREP-examples/Rpkg-mutable) `r emo::ji("package")` is newer and easier
   to grasp. 
 - The [altvecR](https://github.com/s-u/altvecR) `r emo::ji("package")` from Simon Urbanek, which is 
   a *toy package that re-routes ALTREP/ALTVEC methods to R functions for experimentation*. 
 - The actual ALTREP code, mainly lives in the [altrep.c](https://github.com/wch/r-source/blob/trunk/src/main/altrep.c)
   and [Altrep.h](https://github.com/wch/r-source/blob/trunk/src/include/R_ext/Altrep.h) files. 
 - My [altrepisode](https://github.com/romainfrancois/altrepisode). This contains the `stdvec_double` class
   described here and another class that I'll discuss in a follow up post. 
   
